---
title: "**Biostatistics & Epidemiological data analysis using R Final exam**"
subtitle: "Professor: Stefan Konigorski"
author: "Kimia Hariri"
date: "March 5, 2026"
output: 
  pdf_document:
    latex_engine: xelatex
fontsize: 10pt
---

\newpage

# Question 2 - Import, extract and save data

The KiGGS03_06 dataset was downloaded and a new data frame was created with the columns mentioned in the task. The new data frame is as shown below :
```{r , echo=FALSE}
# ================ Question 2-a ================
library(haven)
KiGGS03_06 <- read_sav("KiGGS03_06.sav")

# ================ Question 2-b ================
kiggs_kimia <- data.frame(E070M=KiGGS03_06$E070M , E070V=KiGGS03_06$E070V ,  E072=KiGGS03_06$E072,E074=KiGGS03_06$E074 , sex=KiGGS03_06$sex , age2=KiGGS03_06$age2,schichtz=KiGGS03_06$schichtz , e065z=KiGGS03_06$e065z,HbA1c=KiGGS03_06$HbA1c)

# ================ Question 2-c ================
kiggs_kimia$E070M    <- factor(kiggs_kimia$E070M,    labels = c("yes, daily", "yes, occasionally", "no"))
kiggs_kimia$E070V    <- factor(kiggs_kimia$E070V,    labels = c("yes, daily", "yes, occasionally", "no"))
kiggs_kimia$E072     <- factor(kiggs_kimia$E072,     labels = c("yes, regularly", "yes, from time to time", "no, never"))
kiggs_kimia$E074     <- factor(kiggs_kimia$E074,     labels = c("yes, regularly", "yes, from time to time", "no, never", "has not breastfed"))
kiggs_kimia$sex      <- factor(kiggs_kimia$sex,      labels = c("boys", "girls"))
kiggs_kimia$age2     <- factor(kiggs_kimia$age2,     labels = c("0-1y", "2-3y", "4-5y", "6-7y", "8-9y", "10-11y", "12-13y", "14-15y", "16-17y"))
kiggs_kimia$schichtz <- factor(kiggs_kimia$schichtz, labels = c("low social status", "medium social status", "high social status"))


# kiggs_kimia <- na.omit(kiggs_kimia)
save(kiggs_kimia,file = "kiggs_kimia.Rdata")

head(kiggs_kimia)
```


# Question 3 - Data transformations and data checks

## ================ Question 3-a ================
The format of the variables E070M, E070V, E072, E074 were checked and they were all factors.

```{r}
is.factor(kiggs_kimia$E070M)
is.factor(kiggs_kimia$E070V)
is.factor(kiggs_kimia$E072)
is.factor(kiggs_kimia$E074)
```

The value "has not breastfed" of variable E074 has been set to NA for all children. The E074 column now shows NA for has not breastfed.
```{r, echo=FALSE}

kiggs_kimia$E074[kiggs_kimia$E074=="has not breastfed"] <- NA
head(kiggs_kimia)

```

The levels before setting the value "has not breastfed" of variable E074 to NA for all children.
```{r, echo=FALSE}
levels(kiggs_kimia$E074)
```


The levels after setting the value "has not breastfed" of variable E074 to NA for all children. We do not have the level "has not breastfed" anymore.
```{r, echo=FALSE}
kiggs_kimia$E074 <- droplevels(kiggs_kimia$E074)
levels(kiggs_kimia$E074 )
# We do not have the level "has not breastfed" anymore

```


The new variable burdenS was caluculated. It shows that the higher the value of burdenS, the less exposed children are to smoking, because the levels increase from 'daily' to 'never.'
In other words, a higher level indicates lower smoking exposure.

```{r, echo=FALSE}
burdenS <-as.numeric(kiggs_kimia$E070M)+as.numeric(kiggs_kimia$E070V) +as.numeric(kiggs_kimia$E072) +as.numeric(kiggs_kimia$E074)  
```

## ================ Question 3-b ================
The new variable burdenS was added to the data frame. The updated data frame is as shown below:

```{r, echo=FALSE}

kiggs_kimia$burdenS <-burdenS 
save(kiggs_kimia,file = "kiggs_kimia.Rdata")

head(kiggs_kimia)

```
## ================ Question 3-c ================

Some alternative ways could be instead of just summing the levels we could compute the average of their factor levels.
Another way could be using PCA (principal components), it can be more robust than a simple sum because it can capture important patterns as well.

# Question 4 -  Linear and logistic regression

## ================ Question 4-a ================
I selected age , sex , burdenS from kiggs_kimia data set which I think might be related to HbA1c.

## ================ Question 4-b ================
## Results from the linear regression model using age , sex and burdenS as predictors.
```{r,echo=FALSE}
lm_model <- lm(HbA1c ~ as.numeric(age2) + sex + burdenS,data = kiggs_kimia)
summary(lm_model)
```
## The extracted p-values are shown as below:
```{r,echo=FALSE}
age_p_val <- summary(lm_model)$coefficients[,4][2]
sex_p_val <- summary(lm_model)$coefficients[,4][3]
burdenS_p_val <- summary(lm_model)$coefficients[,4][4]
```
The p-value for the variable age with 3 decimal places: 
```{r,echo=FALSE}
round(age_p_val,3)
```
The p-value for the variable sex with 3 decimal places: 
```{r,echo=FALSE}
round(sex_p_val,3)
```
The p-value for the variable burdenS with 3 decimal places: 
```{r,echo=FALSE}
round(burdenS_p_val,3)
```

The P-value for age is (1.311647e-07) which shows it has a significant effect on HbA1c, for each step increase in age level compared to the reference level, the HbA1c increases 0.009709 unit.

The P-value for sex is (2.145347e-07) which shows it has a significant effect on HbA1c, the HbA1c level decreases by -0.043850 unit in girls compared to boys.

BurdenS does not have significant effect on HbA1c with p-value of 0.434.


## ================ Question 4-c ================
```{r,echo=FALSE}
kiggs_kimia$HbA1c_bi <-ifelse(kiggs_kimia$HbA1c>=5.7 ,1 ,0 ) 
# length(kiggs_kimia$HbA1c_bi[kiggs_kimia$HbA1c_bi==1])
```
A clinically meaningful cutoff of 5.7% was chosen to create the binary HbA1c variable. This threshold is based on the American Diabetes Association (ADA) official standards, as reported by Tucker (2010), which identifies the 5.7%–6.4% range as a 'Category of Increased Risk for Diabetes'. As noted in the report: "The 5.7% threshold was derived from un-published data suggesting that it has the best combination of sensitivity (39%) and specificity (91%) to identify cases of impaired fasting glucose."

## ================ Question 4-d ================

## Below is the summary of the logistic regression model using age , sex and burdenS as predictors.
```{r,echo=FALSE}
log_model <- glm(HbA1c_bi ~ as.numeric(age2) + sex + burdenS, family = binomial(link = "logit"), data = kiggs_kimia)
summary(log_model)
```

```{r,echo=FALSE}
# Report the p-values
age_p_val_log <- summary(log_model)$coefficients[,4][2]
sex_p_val_log <- summary(log_model)$coefficients[,4][3]
burdenS_p_val_log <- summary(log_model)$coefficients[,4][4]
```

## The extracted p-values are shown as shown below:

The p-value for the variable age with 3 decimal places: 
```{r,echo=FALSE}
round(age_p_val_log,3)
```
The p-value for the variable sex with 3 decimal places:
```{r,echo=FALSE}

round(sex_p_val_log,3)
```
The p-value for the variable burdenS with 3 decimal places:
```{r,echo=FALSE}

round(burdenS_p_val_log,3)
```

From the results it can be seen that sex has a p-value of 0.022 which is lower than 0.05, showing a significant association with overall HbA1c.

However, age and burdenS seem to not have a significant association with overall HbA1c level with p-value of 0.702 and 0.859 respectively.

## ================ Question 4-e ================

## The estimated odds ratios for each predictor in the model are as shown below:

```{r,echo=FALSE}

# estimated odds ratios
exp(coef(log_model))

# estimated odds ratios for each predictor in the model
age_OR <- exp(coef(log_model))[2]
sex_OR <- exp(coef(log_model))[3]
burdenS_OR <- exp(coef(log_model))[4]
```

The estimated odds ratios for age with 3 decimal places: 
```{r,echo=FALSE}

round(age_OR,3)
```
The estimated odds ratios for sex with 3 decimal places: 

```{r,echo=FALSE}

round(sex_OR,3)
```
The estimated odds ratios for burdenS with 3 decimal places: 

```{r,echo=FALSE}

round(burdenS_OR,3)
```
## The estimated p-values from significance tests is as shown below:

The p-value for the significance variable sex with 3 decimal places:
```{r,echo=FALSE}
round(summary(log_model)$coefficients[,4][3],3)
```


## The estimates of the 95% confidence interval of the odds ratios for each predictor in the model is :
```{r,echo=FALSE}
exp(confint(log_model))
```

```{r,echo=FALSE}

# # estimated p-values
# summary(log_model)$coefficients[,4]
# 
# # estimated p-values from significance tests 
# summary(log_model)$coefficients[,4][3]
# 
# # estimate of the 95% confidence
# exp(confint(log_model))

```


For the variable age, the Odds Ratio (OR) is 0.9889, meaning that each level increase in age decreases odds of high HbA1c by ~1.1% compared to the age reference level. However the predictors is not statistically associated with high HbA1c in this model because of high p-value (7.021633e-01).

For the variable sex, Odds of high HbA1c decreases by ~27% in girls compared to boys. And the p-value (2.154149e-02) is smaller than 0.05 which shows a strong association.

For burdenS, the Odd Ratio (1.006574) is slightly bigger than 1, suggesting each 1-unit increase in burdenS increases odds by ~0.66%. However the predictors is not statistically associated with high HbA1c in this model because of high p-value (8.589967e-01).

Moreover the CI for age and burdenS include 1 showing they do not have a significant effect on high HbA1c.

The CI for sex exclude 1 ( 0.55799657, 0.95341296 ) showing it reduces odds and it likely have an effect on high HbA1c.

## ================ Question 4-f ================

I think linear regression is more precise because it uses HbA1c as a continuous outcome, meaning it uses all the data and can capture more information. In logistic regression, using a cutoff reduces that information, but it does make the results much easier to interpret for a clinical setting. In my analysis, the linear regression found two associations (age and sex), while the logistic model only found a significant association for sex. This is likely because of the loss of information when we create those categories. Overall, it really depends on the research question and depending on what we want to find, both models can be useful.


# Question 5 - Sample size calculation

## ================ Question 5-a ================

I will use HbA1c as continuous in order to preserve more information.

## ================ Question 5-b ================

I’m choosing an effect size of Cohen’s d=0.32.This comes from a meta-analysis by Cavero-Redondo et al. (2018) (DOI: 10.1007/s40279-018-0861-0) that looked specifically at healthy, non-diabetic people. They found that when you compare an exercise group to a control group, the pooled effect size was 0.32.

I think this is a realistic choice because, in a healthy population (like the kids in KiGGS), their HbA1c is already in a normal range. So we aren't going to see a massive drop like you would in a patient with uncontrolled diabetes, so expecting the effect size of approximately 0.32 is realistic.

## ================ Question 5-c ================

I chose the t-test because the study design is a simple comparison between two independent groups: runners and non-runners. Since the goal is to see if the mean HbA1c differs between these two categories, the t-test is the most direct statistical test to use.

## ================ Question 5-d ================

```{r,echo=FALSE}

# install.packages("pwr")
library(pwr)

effect_size <- 0.32
power <- 0.8
alpha <- 0.05
```
## The results for the minimum necessary sample size calculation is :
```{r,echo=FALSE}
pwr.t.test(d = effect_size, power = power, sig.level = alpha, type = "two.sample", alternative = "two.sided")
```

Based on the results, we would need approximately 154 participants per group (runners 3X/week VS Non-runners) which means in total we need 308 participants

## ================ Question 5-e ================

I do not think this study considers all relevant factors and has some limitations. First, treating running 3x/week as a strict categorical variable may oversimplify the exposure, because some children may run more or less or might stop participating during the study, which could affect the results. Second, the study does not account for other important factors that can influence HbA1c, such as age, BMI, diet, or puberty status. Including these covariates would make the analysis more reliable.


# Question 6 - Meta analysis

## ================ Question 6-a ================

I have chosen to analyze HbA1c as a continuous variable.

## ================ Question 6-b ================

For the exposure variable, I have chosen total sleep duration per day (e065z). which is a continuous variable.

## ================ Question 6-c ================

The population of interest for this study is children and adolescents aged less than 18 years.

## ================ Question 6-d ================

I am choosing to focus on Cross-Sectional Studies because these are much easier to find and perform and they are the most common type of study in public health so there is a much larger pool of research to pull from for this meta-analysis. However, the biggest drawback is that we cannot determine 'temporality.' Since sleep and HbA1c are measured at the same time, we do not know if poor sleep leads to higher HbA1c, or if a child's health issues are what’s causing the poor sleep in the first place.

## ================ Question 6-e ================

## Project 1
Title: Sleep Duration is Correlated with Glycemic Control and Stature in Children with Type 1 Diabetes - a Cross-sectional Study

Author(s): Rayi Kurnia Perwitasari, Yuni Hisbiyah, Nur Rochmah, Muhammad Faizi

Year: 2025

Journal: Clinical Diabetology

DOI: 10.5603/cd.109430

## Project 2 
Title: Sleep in Adolescents and Young Adults with Type 1 Diabetes: Associations with Diabetes Management and Glycemic Control 

Author(s): Sarah S. Jaser, Deborah Ellis 

Year: 2016 

Journal: Health Psychology and Behavioral Medicine 

DOI: 10.1080/21642850.2015.1135293 

## Project 3 
Title: Associations between sleep characteristics and glycemic variability in youth with type 1 diabetes 

Author(s): Necla İpar, Perran Boran, Hatice Ezgi Barış, Mahmut Caner Us, Burcu Aygün, Belma Haliloğlu, Tuğba Gökçe, Ecem Can, Elif Eviz, Neslihan Gökmen İnan, Gül Yeşiltepe Mutlu, Abdullah Bereket, Şükrü Hatun

Year: 2023 

Journal: Sleep Medicine 

DOI: 10.1016/j.sleep.2023.06.018 

## Project 4 
Title: Evaluation of sleep characteristics of children and adolescents with type 1 diabetes mellitus 

Author(s): Renata Aparecida e Silva, Aline De Piano Ganen, Vânia de Fátima Tonetto Fernandes, Nara Michelle de Araújo Evangelista, Carolina Costa Figueiredo, Luciana de Aguiar Pacheco, Guido de Paula Colares Neto 

Year: 2022 

Journal: Revista Paulista de Pediatria 

DOI: 10.1590/1984-0462/2022/40/2020407 

## Project 5 
Title: Impact of Sleep and Activity on Glycemic Control and Quality of Life in Haitian Children and Youth with Type 1 Diabetes 

Author(s): Melanie Babinski, Regina Duperval, Ketly Altenor, Julia von Oettingen 

Year: 2023 

Journal: Pediatric Diabetes 

DOI: 10.1155/2023/4289288 

## Project 6
Title: Sleep Habits in Adolescents with Type 1 Diabetes: Variability in Sleep Duration Linked with Glycemic Control 

Author(s): Niral J. Patel, Kimberly L. Savin, Sachini N. Kahanda, Beth A. Malow, Lauren A. Williams, Gray Lochbihler, Sarah S. Jaser 

Year: 2018 

Journal: Pediatric Diabetes 

DOI: 10.1111/pedi.12689

## Project 7 
Title: Sleep duration and risk of diabetes: Observational and Mendelian randomization studies 

Author(s): Jiao Wang, Man Ki Kwok, Shiu Lun Au Yeung, Albert Martin Li, Hugh Simon Lam, June Yue Yan Leung, Lai Ling Hui, Gabriel Matthew Leung, Catherine Mary Schooling 

Year: 2019 

Journal: Preventive Medicine 

DOI: 10.1016/j.ypmed.2018.11.019 

## Project 8 
Title: Diabetes management mediates the association between sleep duration and glycemic control in youth with type 1 diabetes mellitus

Author(s): Sara S. Frye, Michelle M. Perfect, Graciela E. Silva 

Year: 2019 

Journal: Sleep Medicine 

DOI: 10.1016/j.sleep.2019.01.043 

## Project 9 
Title: The Effect of Sleep Disorders on Blood Sugar in Children With a Type 1 Diabetes Mellitus 

Author(s): Ayeh Shamsadini, Ali Asghar Arabi, Majid Vares Vazirian, Fatemeh Aftabi 

Year: 2023 

Journal: Journal of Pediatrics Review 

DOI: 10.32598/jpr.11.3.1079.1 

## Project 10
Title: The Relationship between Sleep Disturbances and Glycemic Control in Children with Type 1 Diabetes

Author(s): Sana Sarfaraz, Bakhtawar Tahir, Tahir Humaira, Hasnain Faeza

Year: 2025

Journal: Biological and Clinical Sciences Research Journal

DOI: 10.54112/bcsrj.v6i3.1621




(Data formatting and table organization were assisted by the Gemini language model.)


## ================ Question 6-f ================

```{r,include=FALSE}

library(metafor)
library(metadat)
data(dat.bcg, package = "metadat")
# head(dat.bcg)
```

```{r,echo=FALSE}

# 1.
study_perwitasari_2019 <- data.frame(
study = "Perwitasari et al.",
year = 2019,
n = 45,
ri = -0.48
)

# 2.
study_jaser_2016 <- data.frame(
study = "Jaser & Ellis",
year = 2016,
n = 159,
ri = 0.00
)

# 3.
study_ipar_2023 <- data.frame(
  study = "Ipar et al.",
  year = 2023,
  n = 84,
  ri = 0.00
)

# 4.
study_silva_2022 <- data.frame(
  study = "Silva et al.",
  year = 2022,
  n = 86,
  ri = -0.23
)


# 5.
study_babinski_2023 <- data.frame(
  study = "Babinski et al.",
  year = 2023,
  n = 43,
  ri = -0.34
)


# 6
study_patel_2018 <- data.frame(
  study = "Patel et al.",
  year = 2018,
  n = 65,
  ri = -0.10 
)
# 7
wang_2019_hba1c <- data.frame(
  study = "Wang et al.",
  year = 2019,
  n = 3144,
  ri = -0.044    
)


#8
frye_2019 <- data.frame(
  study = "Frye et al.",
  year = 2019,
  n = 111,
  ri = -0.20  
)

# 9
shamsadini_2023 <- data.frame(
  study = "Shamsadini et al.",
  year = 2023,
  n = 200,
  ri = -0.23
)


#10
Sarfaraz_2025 <- data.frame(
  study = "Sarfaraz et al.",
  year = 2025,
  n = 185,
  ri = -0.304
)
```
I reviewed the studies and extracted the correlations between sleep duration and HbA1c level, reported by the researchers. For studies that only reported a standardized beta coefficient, such as Sarfaraz et al. (2025), I converted the value to a correlation (r) using the formula r=0.98(Beta)+0.05(lambda) (Peterson & Brown, 2005). I then created a data frame with all results and used the escalc function, which added the yi and vi columns to calculate the effect size and variance for each study.


I chose the Random-Effects (RE) Model for this meta analysis using the REML (Restricted Maximum Likelihood) method.
Because the chosen studies have extreme diversity. I am  mixing General Population kids (UK) with T1DM kids (Iran, US, etc.) and using different tools (Actigraphy vs. PSQI). A Fixed-Effect model would incorrectly ignore this biological diversity.


(Gemini was used to calculate the correlation coefficients from the reported standardized beta values.)

# The final table is as shown below :
```{r,echo=FALSE}

my_meta_data <- rbind(study_perwitasari_2019,
                      study_jaser_2016,
                      study_ipar_2023,
                      study_silva_2022,
                      study_babinski_2023,
                      study_patel_2018,
                      wang_2019_hba1c,
                      frye_2019,
                      shamsadini_2023,
                      Sarfaraz_2025
                      )

my_meta_data <- escalc(measure="ZCOR", ri=ri, ni=n, data=my_meta_data)
my_meta_data
```

## Results of the Meta-Analysis and Pooled Effect Estimate : 
```{r,echo=FALSE}
res <- rma(yi, vi, data=my_meta_data, method="REML")

# View the results
summary(res)
```
## After converting the pooled Fisher’s Z estimate to the correlation coefficient (r), the results are:
```{r,echo=FALSE}

# Convert the pooled Z back to r
predict(res, transf = transf.ztor)
```


Based on the model results, The pooled Fisher’s Z estimate was -0.1766. After conversion, this corresponds to a correlation of r = -0.17. This is a small-to-moderate negative correlation which  means that as sleep duration increases, HbA1c levels decrease (improve). This result is statistically significant because the p-value is 0.0003, which is below the standard 0.05 threshold. Furthermore, the 95% confidence interval for r ranges from -0.27 to -0.08 and since this range does not include zero, we can be confident that the negative relationship is robust and not due to random chance.

The analysis shows a high level of heterogeneity, with an I^2 value of 72.02%. This percentage tells us that over 70% of the variation in the results is due to real differences between the studies rather than simple measurement error. This is confirmed by the Q-test (p < .0001), which officially rejects the idea that all these studies are measuring the exact same thing in the exact same way. The reason could be because the population of children differs between studies, as well as the varying methods used for sleep measurement (such as objective Actigraphy versus subjective PSQI questionnaires).

(Gemini was used to verify the interpretation of the statistical output and to ensure the findings were described accurately.)

## The formatted forest plot of the results is as shown below : 
```{r,echo=FALSE}


# Create the forest plot
forest(res, 
       transf = transf.ztor,
       slab = my_meta_data$study,
       xlab = "Correlation Coefficient", 
       col = "blue",
       border = "black", 
       main = "Meta-Analysis of Sleep duration and HbA1c")


# Add a vertical line at 0
abline(v = 0, lty = "dashed")


```
Looking at the forest plot, we can see how the 10 studies compare. Some studies, like Perwitasari et al., show a very strong link between more sleep and lower HbA1c, while others, like Jaser & Ellis or Ipar et al., are "neutral" because their horizontal error bars cross the dashed zero line. However, since the entire diamond sits on the left side of the zero line, it proves a statistically significant negative correlation of -0.17. Essentially, this tells us that across all these different groups of children, getting more sleep is consistently associated with better blood sugar levels..

## ================ Question 6-g ================

This meta-analysis has two main weaknesses. First, it is statistically inconsistent because not all researchers reported a direct correlation (ri). For some studies, like Sarfaraz et al. (2025), they only provided a regression coefficient (beta), which forced us to use a math formula to 'guess' or estimate the correlation. Second, it ignores other important factors that affect blood sugar, such as a child’s gender, BMI (weight), or whether they are going through puberty.
